Measuring contrast processing in the visual system using SSVEP (and VEP?)

### 2024-03-29 review.qmd
Adding stuff about sweep

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal. The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

 A signifcant part of research into contrast processing is concerned with how these normalization mechanisms depend on colour, orientation, eye of origin, spatial and temporal frequency, age, and the presence of neurological disorders. The SSVEP has proven to be invaliable in this research because it provides an objective readout of contrast representation at different parts of the visual system and allows us to 'tag' the probe and background separately.

### The development of contrast processing
Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. By varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions measured in single unit studies, or those measuring local field potentials invasively in the cortex. However SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants.

An early use of the contrast response function was to provide an automated estimate of contrast sensitivity, without requiring behavioural responses from the participant. Because SSVEP response at detection threshold are very small, this was achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function to estimate its intercept with the x-axis, which was shown to correspond with psychophysically measured detection thresholds (REFS). This general approach revealed much about the development of visual abilities in infants (Braddick), as well as deficits in neurological disease.

In well-motivated adults, psychophysical measurements remain the gold standard. However, it is hard to obtain reliable psychophysical data from infants, children and adults with neuological disorders. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [Norcia, Tyler, Braddick].

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changes its contrast (or some other property) throughout a trial - for example gradually increasing in contrast or spatial frequency. To avoid hysteresis effects, the sweep was sometimes conducted both up and down in the same experiment [Tyler, Norcia etc].

@@

>>>>

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di Russo, Hillebrand:]. The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

 A signifcant part of research into contrast processing is concerned with how these normalization mechanisms depend on colour, orientation, eye of origin, spatial and temporal frequency, age, and the presence of neurological disorders [refs: Chen and Foley, Petrov etc Carandini, Binocular normalization / suppression (Baker/Meese but also e.g. Heeger), Tadin, Winawer, Wade, Porciatti, Tsai...]. The SSVEP has proven to be invaliable in this research because it provides an objective readout of contrast representation at different parts of the visual system and allows us to 'tag' the probe and background separately.

### The development of contrast processing
Because it provides a direct read-out of neural population activity, the SSVEP signal can reveal key features of neural signal transduction. For example, by varying the peak stimulus contrast parametrically, a 'contrast vs response' function can be measured - where the 'response' is typically defined as the amplitude of the SSVEP frequency component at the stimulus frequency or a low multiple thereof. This corresponds closely to similar functions measured in single unit studies, or those measuring local field potentials invasively in the cortex[@ShapleyVictor etc]. However the SSVEP has the advantage that it is non-invasive, and so can be measured in awake, behaving human participants.

An early use of the contrast response function was to provide an automated estimate of contrast sensitivity, without requiring behavioural responses from the participant. Because SSVEP response at detection threshold are very small, this was achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function to estimate its intercept with the x-axis, which was shown to correspond with psychophysically measured detection thresholds [(REFS)]. This general approach revealed much about the development of visual abilities in infants [@braddick], as well as deficits in neurological disease.

In well-motivated adults, psychophysical measurements of contrast sensitivity remain the gold standard. However, it is hard to obtain reliable psychophysical data from infants, children and adults with neuological disorders. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [Norcia, Tyler, Braddick, Porciatti, Tsai...].

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, or some other property, throughout a trial - for example gradually increasing in spatial frequency [Tyler, Nakayama etc 1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [Tyler, Norcia etc: 1985, 1990 etc].

In general, SSVEP measurements of infant vision have revealed that the development of spatial acuity for both achromatic and chromatic contrast as well as stereoscopic depth perception was faster than had been supposed previously based on behavioural measures [@DobsonTeller78, @NorciaTyler85, @NorciaGerhard 15]

@@

As well as measuing contrast thresholds for a stimulus with a fixed spatiotemporal frequency, this technique can also be used to assess spatial frequency thresholds [ REFS - Almoqbel etc.]. Because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in young subjects [Ridder2004 Diva vs Enfant, Hoffman, Bach and Farmer].
